---
title: "Community Suitability Modelling of Quebec Forests with HMSC"
subtitle: "Reproducible Analysis Code for: Beyond Moving a Single Species: Using Community Suitability to Inform Climate Adaptation in the Boreal-Temperate Ecotone"
author: "João Paulo C. Liz¹, Nelson Thiffault², David Voyer¹, Loïc D’Orangeville¹, Nicholas C. Coops³ and Alexis Achim¹\n¹Faculty of Forestry, Geography and Geomatics, Laval University, Quebec, Canada
²Canadian Forest Service, Natural Resources Canada, Quebec, Canada
³Department of Forest Resources Management, University of British Columbia, Vancouver, Canada
"
date: "`r paste(month.name[as.integer(format(Sys.Date(), '%m'))], format(Sys.Date(), '%d, %Y'))`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    code_folding: show
    theme: readable
    df_print: kable
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
fontsize: 11pt
geometry: margin=1in
---

```{r global-options, include = FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  message    = FALSE,
  warning    = FALSE,
  fig.align  = "center",
  fig.width  = 7,
  fig.height = 5,
  dpi        = 150,
  linewidth  = 80
)
options(scipen = 999)

# Wrap long source lines in the PDF output so code never overflows the page
# margin (HTML output already wraps/scrolls, so this only affects the PDF).
hook_source <- knitr::knit_hooks$get("source")
knitr::knit_hooks$set(source = function(x, options) {
  if (knitr::is_latex_output() && !is.null(n <- options$linewidth)) {
    x <- knitr:::split_lines(x)
    if (any(nchar(x) > n)) x <- strwrap(x, width = n, exdent = 2)
    x <- paste(x, collapse = "\n")
  }
  hook_source(x, options)
})
```

# Overview

This document is the reproducible companion code for the manuscript *"Beyond
Single Species: Using Community Suitability to Inform Climate Adaptation in
the Boreal-Temperate Ecotone"*. It walks through every analytical step
described in the Methods section, in the order they are described there:

| Section below                                   | Manuscript section                       | Original script                                |
|--------------------------------------------------|-------------------------------------------|--------------------------------------------------|
| 2. Study area and sampling design                | Section 2 (Study Area)                    | `000_study_area_map.R`                          |
| 3. Forest inventory data and predictors           | 3.1 Data                                  | `001_fit_hmsc_model.R` (data section)           |
| 4. HMSC model specification and fitting           | 3.2.1--3.2.2 Modeling                     | `001_fit_hmsc_model.R` (model section)          |
| 5. MCMC convergence diagnostics                   | 3.2.2 / Figure A3                          | `001_fit_hmsc_model.R`                          |
| 6. Explanatory and predictive model performance   | 3.2.2 / Table 2                            | `001_fit_hmsc_model.R`, `008_cv_predictive_power.R`, `009_cv_results_table_plots.R` |
| 7. Variance partitioning                          | 3.2.3 / Figure 1                           | `001_fit_hmsc_model.R`                          |
| 8. Species environmental niches (beta)            | 3.2.3 / Figure 4                           | `001_fit_hmsc_model.R`                          |
| 9. Species associations (Omega)                   | 3.2.3 / Figure A4-A5                       | `001_fit_hmsc_model.R`                          |
| 10. Future climate scenarios and predictions      | 3.1.3, 3.2.3                                | `002_predict_future_scenarios.R`                |
| 11. Community Suitability Index (CSI)             | 3.3                                         | `003_compute_csi_from_draws.R`                  |
| 12. CSI trajectories under climate change         | 4.3 / Figure 3, Figure A7                   | `003a_plot_csi_trajectories.R`                  |
| 13. Appendix: draws subsampling for the TCAMS app | 3.5 (TCAMS tool)                            | `004_prepare_app_draws_subsample.R`             |

## How to use this document

Several steps in this pipeline are **computationally heavy** -- most notably
fitting the HMSC model itself (four MCMC chains, 150,000 iterations each),
the four-fold cross-validation used to assess predictive power (which refits
the model once per fold), and generating posterior predictions for the 12
future climate scenarios. Re-running these from scratch is **not required**
to reproduce the manuscript's tables and figures, since every heavy step
saves its output to disk, and this document loads the saved object right
after showing the code that produced it.

A single switch controls this behaviour:

```{r run-heavy-flag}
# Set to TRUE only if you want to literally re-run the multi-hour steps
# (model fitting, cross-validation, future-climate prediction) from raw data.
# Leave FALSE to reproduce all tables/figures from the saved outputs that
# accompany this submission.
RUN_HEAVY <- FALSE
```

Every heavy chunk below is guarded by `eval = RUN_HEAVY`, and is immediately
followed by a chunk guarded by `eval = !RUN_HEAVY` that loads the
corresponding pre-computed object. Either way, by the end of each section the
same R objects are available in the environment, so downstream code and
figures render identically.

# 1. Environment and Path Configuration

```{r setup-libs}
# --- Library path -----------------------------------------------------------
# Edit this if reproducing on a different machine; it must point to the
# folder containing the Hmsc, coda, BayesLogit, sf, tidyverse, etc. packages.
.libPaths("H:/R_lib")
options(repos = c(CRAN = "https://cloud.r-project.org"))

pkgs <- c("tidyverse", "sf", "Hmsc", "coda", "BayesLogit", "knitr",
          "kableExtra", "ggcorrplot", "igraph", "ggraph", "scales",
          "viridis", "ggrepel", "rnaturalearth", "rnaturalearthdata",
          "ggspatial", "patchwork", "abind")
for (p in pkgs) {
  if (!requireNamespace(p, quietly = TRUE)) install.packages(p, lib = "H:/R_lib")
}

suppressPackageStartupMessages({
  library(tidyverse)
  library(sf)
  library(BayesLogit)   # required internally by Hmsc's parallel workers
  library(Hmsc)
  library(coda)
  library(knitr)
  library(kableExtra)
})
```

```{r setup-paths}
# --- Project folders ---------------------------------------------------------
# All paths below are relative to a single base directory. Edit BASE_DIR to
# point to the "PEP_QC" project root on your own machine; everything else is
# derived from it and does not need to be touched.
BASE_DIR    <- "G:/Thesis/3rdChapter/PEP_QC"
DATA_DIR    <- file.path(BASE_DIR, "databases")
RESULTS_DIR <- file.path(BASE_DIR, "results", "undisturbed")
MODEL_DIR   <- file.path(RESULTS_DIR, "hmsc")
FIGS_DIR    <- file.path(RESULTS_DIR, "hmsc_figures")
MAPS_DIR    <- file.path(BASE_DIR, "results", "maps")
OUTPUTS_DIR <- file.path(RESULTS_DIR, "outputs")
DRAWS_DIR   <- file.path(BASE_DIR, "posterior_draws1")

MODEL_FILE  <- file.path(MODEL_DIR, "hmsc_presence_model.Rdata")
CV_FILE     <- file.path(MODEL_DIR, "hmsc_presence_model_CV.Rdata")

for (d in c(FIGS_DIR, MAPS_DIR, OUTPUTS_DIR)) {
  dir.create(d, recursive = TRUE, showWarnings = FALSE)
}
```

```{r species-dictionary}
# --- Species code dictionary -------------------------------------------------
# Maps the 3-letter Quebec forest-inventory species codes (e.g. "SAB") used
# throughout the raw data and objects to full scientific names and to the
# short "Genus.species" labels used in figures (e.g. "Abi.bals").
species_map <- tibble::tribble(
  ~code, ~french_name, ~scientific_name,
  "AEH", "Marronnier d'Inde", "Aesculus hippocastanum",
  "AME", "Amelanchier", "Amelanchier spp.",
  "AUC", "Aulne crispe", "Alnus crispa",
  "AUR", "Aulne rugueux", "Alnus incana",
  "BOG", "Bouleau gris", "Betula populifolia",
  "BOJ", "Bouleau jaune", "Betula alleghaniensis",
  "BOP", "Bouleau a papier", "Betula papyrifera",
  "CAC", "Caryer cordiforme", "Carya cordiformis",
  "CAF", "Caryer ovale", "Carya ovata",
  "CAR", "Charme de Caroline", "Carpinus caroliniana",
  "CET", "Cerisier tardif", "Prunus serotina",
  "CHB", "Chene blanc", "Quercus alba",
  "CHE", "Chene bicolore", "Quercus bicolor",
  "CHG", "Chene a gros fruits", "Quercus macrocarpa",
  "CHR", "Chene rouge", "Quercus rubra",
  "EPB", "Epinette blanche", "Picea glauca",
  "EPN", "Epinette noire", "Picea mariana",
  "EPO", "Epinette de Norvege", "Picea abies",
  "EPR", "Epinette rouge", "Picea rubens",
  "ERA", "Erable argente", "Acer saccharinum",
  "ERB", "Erable de Norvege", "Acer platanoides",
  "ERE", "Erable a epis", "Acer spicatum",
  "ERG", "Erable a Giguere", "Acer negundo",
  "ERN", "Erable noir", "Acer nigrum",
  "ERP", "Erable de Pennsylvanie", "Acer pensylvanicum",
  "ERR", "Erable rouge", "Acer rubrum",
  "ERS", "Erable a sucre", "Acer saccharum",
  "FRA", "Frene d'Amerique", "Fraxinus americana",
  "FRN", "Frene noir", "Fraxinus nigra",
  "FRP", "Frene de Pennsylvanie", "Fraxinus pennsylvanica",
  "HEG", "Hetre a grandes feuilles", "Fagus grandifolia",
  "MEL", "Meleze laricin", "Larix laricina",
  "NOC", "Noyer cendre", "Juglans cinerea",
  "ORA", "Orme d'Amerique", "Ulmus americana",
  "ORR", "Orme rouge", "Ulmus rubra",
  "ORT", "Orme de Thomas", "Ulmus thomasii",
  "OSV", "Ostryer de Virginie", "Ostrya virginiana",
  "PEB", "Peuplier baumier", "Populus balsamifera",
  "PEG", "Peuplier a grandes dents", "Populus grandidentata",
  "PET", "Peuplier faux-tremble", "Populus tremuloides",
  "PIB", "Pin blanc", "Pinus strobus",
  "PIG", "Pin gris", "Pinus banksiana",
  "PIR", "Pin rouge", "Pinus resinosa",
  "PIS", "Pin sylvestre", "Pinus sylvestris",
  "PRU", "Pruche de l'Est", "Tsuga canadensis",
  "PRP", "Cerisier de Pennsylvanie", "Prunus pensylvanica",
  "SAB", "Sapin baumier", "Abies balsamea",
  "SOA", "Sorbier d'Amerique", "Sorbus americana",
  "SOD", "Sorbier des montagnes", "Sorbus decora",
  "THO", "Thuya occidental", "Thuja occidentalis",
  "TIL", "Tilleul d'Amerique", "Tilia americana"
) %>%
  mutate(
    original_name_in_data = paste0(code, "_n"),
    pretty_name = sapply(scientific_name, function(sci) {
      parts   <- strsplit(sci, " ")[[1]]
      genus   <- substr(parts[1], 1, 3)
      species <- if (length(parts) > 1) substr(parts[2], 1, 4) else ""
      paste0(stringr::str_to_title(genus), ".", species)
    })
  )

knitr::kable(species_map %>% select(code, scientific_name, pretty_name),
             caption = "Species code dictionary used throughout this document.")
```

# 2. Study Area and Sampling Design (Figure 1)

The study area covers Quebec's temperate and boreal forest regions south of
the 52nd parallel. The map below shows the 5,572 permanent sample plots used
in the final model together with the ten bioclimatic domains of Quebec
(Saucier et al., 2009), which give visual context to the boreal-temperate
ecotone referenced throughout the manuscript.

```{r figure1-map, fig.cap = "Study area: permanent sample plots (coloured by Mean Annual Temperature) over Quebec's bioclimatic domains, ordered from temperate hardwood (south) to forest tundra (north)."}
suppressPackageStartupMessages({
  library(rnaturalearth); library(rnaturalearthdata)
  library(ggspatial); library(patchwork); library(scales); library(viridis)
  library(ggrepel)
})

# --- Permanent sample plots used in the fitted model -------------------------
# We restrict to the exact set of sites that entered the fitted model, so
# that "n" in the figure caption always matches the model's sample size,
# even if the raw inventory CSV contains additional (later-dropped) plots.
model_env <- new.env(parent = emptyenv())
load(MODEL_FILE, envir = model_env)
model_sites <- rownames(model_env$m1$Y)

plots_sf <- read_csv(file.path(DATA_DIR, "undisturbed_plots_sf_env_clim_4th_inv.csv"),
                      show_col_types = FALSE) %>%
  dplyr::select(plot_id, latitude, longitude, MAT, MAP) %>%
  filter(!is.na(longitude), !is.na(latitude),
         as.character(plot_id) %in% model_sites) %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326)

cat("Plots in fitted model:", length(model_sites),
    "| Plots shown on map:", nrow(plots_sf), "\n")

# --- Bioclimatic domains (Saucier et al. 2009) -------------------------------
# N3_DOM_BIO layer of Quebec's official ecological classification geodatabase:
# the ten classic bioclimatic domains, ordered south (temperate hardwood) to
# north (forest tundra); directly matches the community codes used in Table 1
# of the manuscript (e.g. domain 3 = community FE3, domain 6 = community RE2).
gdb_zip <- file.path(BASE_DIR, "MAT_QUEBEC_VEG", "CLASSI_ECO_QC_GDB.zip")
gdb_vsizip <- paste0("/vsizip/", gdb_zip, "/CLASSI_ECO_QC_GDB/CLASSI_ECO_QC.gdb")

dom_bio_raw <- st_read(gdb_vsizip, layer = "N3_DOM_BIO", quiet = TRUE)
if (is.na(st_crs(dom_bio_raw))) st_crs(dom_bio_raw) <- 32198  # NAD83 / Quebec Lambert
dom_bio_raw <- st_make_valid(dom_bio_raw)

domain_labels <- c(
  "Erabliere a caryer cordiforme" = "1. Sugar maple - bitternut hickory",
  "Erabliere a tilleul"           = "2. Sugar maple - basswood",
  "Erabliere a bouleau jaune"     = "3. Sugar maple - yellow birch",
  "Sapiniere a bouleau jaune"     = "4. Balsam fir - yellow birch",
  "Sapiniere a bouleau a papier"  = "5. Balsam fir - paper birch",
  "Pessiere a mousses"            = "6. Black spruce - moss",
  "Pessiere a lichens"            = "7. Black spruce - lichen",
  "Toundra forestiere"            = "8. Forest tundra",
  "Toundra a arbustes dresses"    = "9. Shrub tundra (erect)",
  "Toundra a arbustes prostres"   = "10. Shrub tundra (prostrate)"
)
# NOM_DB is stored with accented French characters; normalise before recoding.
nom_ascii <- iconv(dom_bio_raw$NOM_DB, from = "UTF-8", to = "ASCII//TRANSLIT")
dom_bio_raw$domain_en <- factor(dplyr::recode(nom_ascii, !!!domain_labels),
                                 levels = unname(domain_labels))

domain_colors <- c(
  "1. Sugar maple - bitternut hickory" = "#1B5E20",
  "2. Sugar maple - basswood"          = "#33691E",
  "3. Sugar maple - yellow birch"      = "#7CB342",
  "4. Balsam fir - yellow birch"       = "#AED581",
  "5. Balsam fir - paper birch"        = "#4FC3F7",
  "6. Black spruce - moss"             = "#0288D1",
  "7. Black spruce - lichen"           = "#01579B",
  "8. Forest tundra"                   = "#78909C",
  "9. Shrub tundra (erect)"            = "#B0BEC5",
  "10. Shrub tundra (prostrate)"       = "#CFD8DC"
)

# --- Reference boundaries -----------------------------------------------------
LAMBERT_QC <- 32198
all_states <- sf::st_as_sf(rnaturalearthdata::states50) %>% st_make_valid()
provinces  <- all_states %>% filter(admin == "Canada")
us_states  <- all_states %>% filter(admin == "United States of America")
world      <- ne_countries(scale = "medium", returnclass = "sf") %>% st_make_valid()
quebec_sf  <- provinces %>% filter(name == "Quebec") %>% st_make_valid()
context_sf <- bind_rows(
  provinces %>% filter(name %in% c("Ontario", "New Brunswick", "Nova Scotia",
                                    "Newfoundland and Labrador", "Prince Edward Island")),
  us_states %>% filter(name %in% c("Maine", "Vermont", "New Hampshire",
                                    "New York", "Massachusetts"))
) %>% st_make_valid()

bbox_wgs <- st_bbox(c(xmin = -80, xmax = -57, ymin = 44.5, ymax = 53), crs = st_crs(4326))
bbox_lam <- st_bbox(st_transform(st_as_sfc(bbox_wgs), LAMBERT_QC))
clip_lam       <- st_as_sfc(bbox_lam) %>% st_buffer(500000) %>% st_make_valid()
clip_lam_tight <- st_as_sfc(bbox_lam) %>% st_buffer(50000)  %>% st_make_valid()

safe_project <- function(x, crs, clip) {
  x %>% st_transform(crs) %>% st_make_valid() %>%
    st_intersection(clip) %>% st_make_valid()
}
world_p   <- safe_project(world,      LAMBERT_QC, clip_lam)
context_p <- safe_project(context_sf, LAMBERT_QC, clip_lam)
dom_bio_p <- safe_project(dom_bio_raw, LAMBERT_QC, clip_lam_tight)
plots_p   <- st_transform(plots_sf, LAMBERT_QC)

parallel_52 <- st_linestring(matrix(c(-82, 52, -56, 52), ncol = 2, byrow = TRUE)) %>%
  st_sfc(crs = 4326) %>% st_sf() %>% st_transform(LAMBERT_QC)
label_52N <- st_coordinates(st_transform(st_sfc(st_point(c(-73, 52.1)), crs = 4326), LAMBERT_QC))

cities <- tribble(
  ~name,           ~lon,      ~lat,
  "Montreal",      -73.5673,  45.5017,
  "Quebec City",   -71.2080,  46.8139,
  "Sherbrooke",    -71.8989,  45.4042,
  "Rouyn-Noranda", -79.0210,  48.2360,
  "Saguenay",      -71.0596,  48.4281,
  "Sept-Iles",     -66.3821,  50.2101,
  "Gaspe",         -64.4859,  48.8339
) %>% st_as_sf(coords = c("lon", "lat"), crs = 4326) %>% st_transform(LAMBERT_QC)

theme_map <- theme_void(base_size = 11) +
  theme(panel.background = element_rect(fill = "#B8D4E8", colour = NA),
        panel.border     = element_rect(fill = NA, colour = "grey35", linewidth = 0.7),
        legend.position  = "right",
        plot.title       = element_text(size = 12, face = "bold", hjust = 0.5))

ggplot() +
  geom_sf(data = world_p,   fill = "#EAE8E2", colour = "grey65",  linewidth = 0.10) +
  geom_sf(data = context_p, fill = "#D8D3CB", colour = "#888888", linewidth = 0.35) +
  geom_sf(data = dom_bio_p, aes(fill = domain_en), colour = "grey30", linewidth = 0.12) +
  scale_fill_manual(name = "Bioclimatic domain\n(Saucier et al. 2009)",
                     values = domain_colors, drop = TRUE) +
  geom_sf(data = parallel_52, colour = "#CC3333", linewidth = 0.6, linetype = "dashed") +
  annotate("text", x = label_52N[1], y = label_52N[2], label = "52\u00b0N (study limit)",
           colour = "#CC3333", size = 3, fontface = "italic") +
  geom_sf(data = plots_p, aes(colour = MAT), size = 0.55, alpha = 0.85, shape = 16) +
  scale_colour_viridis_c(name = "Plot MAT (\u00b0C)", option = "plasma") +
  geom_sf(data = cities, shape = 18, size = 2, colour = "black") +
  geom_label_repel(data = cities, aes(geometry = geometry, label = name),
                    stat = "sf_coordinates", size = 2.6, seed = 1304,
                    max.overlaps = Inf, force = 8, box.padding = 0.6) +
  coord_sf(xlim = c(bbox_lam["xmin"], bbox_lam["xmax"]),
           ylim = c(bbox_lam["ymin"], bbox_lam["ymax"]),
           expand = FALSE, crs = LAMBERT_QC, datum = st_crs(4326)) +
  annotation_scale(location = "tl", width_hint = 0.22) +
  annotation_north_arrow(location = "tl", which_north = "true",
                          pad_y = unit(0.35, "in")) +
  labs(title = "Study Area: Permanent Sample Plots and Bioclimatic Domains",
       subtitle = paste0("Quebec, Canada | n = ", format(nrow(plots_sf), big.mark = ","),
                          " undisturbed plots | below 52\u00b0N latitude")) +
  theme_map
```

# 3. Forest Inventory Data and Predictors (Section 3.1)

We use plot-level data from the 4th decennial forest inventory of Quebec's
permanent sample plot network, restricted to plots without recorded
anthropogenic disturbance. Species with fewer than 5% occupied plots are
dropped, and only plots with complete environmental data are retained --
yielding the 5,572-plot, 13-species dataset reported in the manuscript.

```{r data-prep}
plots_sf_env_clim <- read_csv(file.path(DATA_DIR, "undisturbed_plots_sf_env_clim_4th_inv.csv"),
                               show_col_types = FALSE)

# Climatic, edaphic and topographic predictors (Section 3.1.2)
env_predictors <- c("MAP", "MAT", "pH", "Clay", "OM", "rtp", "twi",
                     "slope", "drainage_class")
plots_sf_env_clim$drainage_class <- as.factor(plots_sf_env_clim$drainage_class)

species_abun_vars <- plots_sf_env_clim %>%
  select(ends_with("_n") & !starts_with("NA")) %>%
  colnames()

Y_full <- plots_sf_env_clim %>%
  dplyr::select(plot_id, all_of(species_abun_vars)) %>%
  mutate(across(all_of(species_abun_vars), ~ as.integer(replace_na(., 0)))) %>%
  as.data.frame() %>%
  column_to_rownames(var = "plot_id") %>%
  as.matrix()

# Drop species present in fewer than 5% of plots
prevalence           <- colSums(Y_full > 0, na.rm = TRUE)
prevalence_threshold <- nrow(Y_full) * 0.05
species_to_keep      <- names(prevalence[prevalence >= prevalence_threshold])
Y <- Y_full[, species_to_keep]

XData <- plots_sf_env_clim %>%
  dplyr::select(plot_id, all_of(env_predictors)) %>%
  mutate_if(is.character, as.factor) %>%
  as.data.frame() %>%
  column_to_rownames(var = "plot_id")

# Align plots with complete environmental data and non-zero occupancy
common_plots <- intersect(rownames(Y), rownames(na.omit(XData)))
Y      <- Y[common_plots, ]
XData  <- XData[common_plots, ]
Y      <- Y[, colSums(Y) > 0]
Y[Y > 0] <- 1  # convert abundance to presence-absence for the probit model

cat("Final data dimensions -- Y (species matrix):", nrow(Y), "plots x", ncol(Y), "species\n")
cat("Final data dimensions -- XData (covariates): ", nrow(XData), "plots x", ncol(XData), "covariates\n")
```

# 4. HMSC Model Specification and Fitting (Sections 3.2.1-3.2.2)

We fit a Hierarchical Modelling of Species Communities (HMSC) probit joint
species distribution model (Ovaskainen et al. 2017; Tikhonov et al. 2020),
with a site-level random effect capturing residual (biotic + unmeasured)
co-occurrence structure.

```{r hmsc-specify, eval = RUN_HEAVY}
XFormula <- as.formula(paste("~", paste(env_predictors, collapse = " + ")))

studyDesign <- data.frame(site = as.factor(rownames(Y)))
rownames(studyDesign) <- rownames(Y)
rL_site <- HmscRandomLevel(units = studyDesign$site)

mod <- Hmsc(
  Y = Y, XData = XData, XFormula = XFormula,
  distr = "probit",
  studyDesign = studyDesign,
  ranLevels = list(site = rL_site)
)
```

```{r hmsc-fit, eval = RUN_HEAVY}
# Four MCMC chains, 150,000 iterations each (50,000 burn-in + 100,000 sampling
# thinned by 100 -> 1,000 posterior samples per chain, 4,000 in total).
# This step alone takes on the order of an hour on a modern multi-core machine.
nChains <- 4
thin    <- 100
samples <- 1000

set.seed(1304)
m1 <- sampleMcmc(
  mod, samples = samples, thin = thin,
  adaptNf   = rep(ceiling(0.4 * samples * thin), 1),
  transient = ceiling(0.5 * samples * thin),
  nChains = nChains, nParallel = nChains,
  initPar = "fixed effects"
)

save(m1, file = MODEL_FILE)
```

```{r hmsc-load, eval = !RUN_HEAVY}
# --- Load the already-fitted model instead of re-running the MCMC ----------
load(MODEL_FILE)  # loads m1
```

```{r hmsc-summary}
nChains <- length(m1$postList)
cat("Fitted HMSC model summary\n")
cat("  chains   :", nChains, "\n")
cat("  samples  :", m1$samples, "per chain (thin =", m1$thin, ", transient =", m1$transient, ")\n")
cat("  sites    :", m1$ny, "\n")
cat("  species  :", m1$ns, "\n")
cat("  distr    : probit\n")
```

# 5. MCMC Convergence Diagnostics (Figure A3)

We assess convergence of the species-environmental response parameters
(*Beta*) using the Effective Sample Size (ESS) and the Gelman-Rubin Potential
Scale Reduction Factor (PSRF). *Note: convergence is checked here for the Beta
parameters specifically; the residual association matrix (Omega) and
random-effect hyperparameters are not separately diagnosed.*

```{r convergence-diagnostics, fig.cap = "MCMC convergence diagnostics for the species-environmental response parameters (Beta)."}
mpost <- convertToCodaObject(m1)

ess.beta  <- effectiveSize(mpost$Beta)
psrf.beta <- gelman.diag(mpost$Beta, multivariate = FALSE)$psrf[, "Point est."]

cat("Beta ESS  : median =", round(median(ess.beta), 0),
    "| min =", round(min(ess.beta), 0), "\n")
cat("Beta PSRF : median =", round(median(psrf.beta), 3),
    "| max =", round(max(psrf.beta), 3), "\n")

par(mfrow = c(1, 2))
hist(ess.beta, breaks = 30, col = "skyblue", main = "Effective Sample Size (Beta)",
     xlab = "ESS")
abline(v = 1000, col = "red", lty = 2, lwd = 2)
hist(psrf.beta, breaks = 30, col = "lightgreen", main = "Gelman-Rubin PSRF (Beta)",
     xlab = "PSRF (point estimate)")
abline(v = 1.1, col = "red", lty = 2, lwd = 2)
par(mfrow = c(1, 1))
```

# 6. Explanatory and Predictive Model Performance (Table 2)

## 6.1 Explanatory power (in-sample fit)

```{r explanatory-fit}
preds_explanatory <- computePredictedValues(m1)
MF <- evaluateModelFit(hM = m1, predY = preds_explanatory)

cat(sprintf("Mean explanatory AUC     = %.3f (SD = %.3f)\n", mean(MF$AUC), sd(MF$AUC)))
cat(sprintf("Mean explanatory Tjur R2 = %.3f (SD = %.3f)\n", mean(MF$TjurR2), sd(MF$TjurR2)))
```

## 6.2 Predictive power (4-fold cross-validation)

Explanatory metrics above are computed in-sample and are therefore
optimistic. We additionally assess **predictive power** through four-fold
cross-validation (`createPartition`), following the same protocol used by
Tikhonov et al. (2020, Appendix S2). Each fold refits the model with the
same MCMC settings, so this step takes roughly four times as long as the
original fit (about 4.8 hours on the machine used for this analysis).

```{r cv-run, eval = RUN_HEAVY}
set.seed(1304)
partition <- createPartition(m1, nfolds = 4, column = "site")

predsCV <- computePredictedValues(m1, partition = partition, nParallel = nChains)
MFCV <- evaluateModelFit(hM = m1, predY = predsCV)

save(predsCV, MFCV, partition, file = CV_FILE)
```

```{r cv-load, eval = !RUN_HEAVY}
# --- Load pre-computed cross-validation results -----------------------------
cv_env <- new.env(parent = emptyenv())
load(CV_FILE, envir = cv_env)  # loads predsCV, MFCV, partition
MFCV <- cv_env$MFCV
```

```{r cv-table}
model_fit_df <- tibble(
  Species            = colnames(m1$Y),
  Prevalence         = colSums(m1$Y > 0),
  Explanatory_AUC    = MF$AUC,
  Predictive_AUC     = MFCV$AUC,
  Explanatory_TjurR2 = MF$TjurR2,
  Predictive_TjurR2  = MFCV$TjurR2,
  Explanatory_RMSE   = MF$RMSE,
  Predictive_RMSE    = MFCV$RMSE
) %>%
  left_join(species_map, by = c("Species" = "original_name_in_data")) %>%
  arrange(desc(Explanatory_TjurR2)) %>%
  dplyr::select(pretty_name, Prevalence, Explanatory_AUC, Predictive_AUC,
                Explanatory_TjurR2, Predictive_TjurR2)

cat(sprintf("Mean predictive AUC     = %.3f (SD = %.3f)\n",
            mean(MFCV$AUC), sd(MFCV$AUC)))
cat(sprintf("Mean predictive Tjur R2 = %.3f (SD = %.3f)\n",
            mean(MFCV$TjurR2), sd(MFCV$TjurR2)))

knitr::kable(model_fit_df, digits = 3,
             col.names = c("Species", "Prevalence", "Explanatory AUC",
                           "Predictive AUC", "Explanatory Tjur R2", "Predictive Tjur R2"),
             caption = "Table 2. Explanatory (in-sample) and predictive (4-fold cross-validation) model performance for each of the 13 tree species.")
```

```{r cv-plot, fig.cap = "Explanatory vs. predictive AUC per species. Points below the dashed 1:1 line indicate the degree of overfitting."}
ggplot(model_fit_df, aes(x = Explanatory_AUC, y = Predictive_AUC)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50") +
  geom_point(color = "darkblue", size = 3, alpha = 0.7) +
  ggrepel::geom_text_repel(aes(label = pretty_name), size = 3) +
  coord_equal(xlim = c(0.6, 1), ylim = c(0.6, 1)) +
  labs(x = "Explanatory AUC (in-sample)",
       y = "Predictive AUC (4-fold cross-validation)") +
  theme_bw()
```

# 7. Variance Partitioning (Figure 1 of the manuscript)

```{r variance-partitioning, fig.cap = "Variance partitioning of species occurrence into fixed-effect predictor groups and the random site effect."}
VP <- computeVariancePartitioning(m1)

vp_df <- VP$vals %>%
  as_tibble(rownames = "variable") %>%
  pivot_longer(cols = -variable, names_to = "Species", values_to = "value") %>%
  left_join(species_map, by = c("Species" = "original_name_in_data"))

variable_importance <- vp_df %>%
  group_by(variable) %>%
  summarise(mean_value = mean(value, na.rm = TRUE)) %>%
  arrange(desc(mean_value))

knitr::kable(variable_importance, digits = 3,
             caption = "Mean proportion of explained variance attributed to each predictor group / the random site effect, averaged across species.")

ggplot(vp_df, aes(x = value, y = pretty_name, fill = variable)) +
  geom_bar(stat = "identity", position = "stack") +
  scale_x_continuous(labels = scales::percent) +
  labs(x = "Proportion of Explained Variance", y = "Species", fill = "Predictor group") +
  theme_classic() +
  theme(legend.position = "bottom")
```

# 8. Species Environmental Niches (Beta Coefficients, Figure 4)

```{r species-niches, fig.cap = "Posterior mean Beta coefficients with >=95% posterior support (species x environmental predictor)."}
postBeta <- getPostEstimate(m1, parName = "Beta")
predictor_names <- m1$covNames

means_df <- postBeta$mean %>% as_tibble() %>%
  mutate(variable = predictor_names, .before = 1) %>%
  pivot_longer(cols = -variable, names_to = "Species", values_to = "Mean")

supported_df <- postBeta$support %>% as_tibble() %>%
  mutate(variable = predictor_names, .before = 1) %>%
  pivot_longer(cols = -variable, names_to = "Species", values_to = "Support") %>%
  left_join(means_df, by = c("variable", "Species")) %>%
  dplyr::filter((Support > 0.95 | Support < 0.05), variable != "(Intercept)") %>%
  mutate(sign = ifelse(Mean > 0, "+", "-")) %>%
  left_join(species_map, by = c("Species" = "original_name_in_data")) %>%
  dplyr::filter(!is.na(pretty_name))

ggplot(supported_df, aes(x = variable, y = pretty_name, fill = Mean, color = sign)) +
  geom_tile(linewidth = 0.5) +
  scale_fill_gradient2(low = "darkblue", mid = "grey90", high = "darkred",
                        midpoint = 0, name = "Mean Beta") +
  scale_color_manual(values = c("+" = "red", "-" = "blue")) +
  guides(color = "none") +
  labs(x = "Environmental predictor", y = "Species") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
        legend.position = "bottom")
```

# 9. Species Associations (Residual Correlation Matrix Omega, Figure A4-A5)

```{r species-associations, fig.cap = "Significant residual species co-occurrences (posterior support > 95%), after accounting for shared environmental responses."}
OmegaCor <- computeAssociations(m1)
cor_matrix_mean <- OmegaCor[[1]]$mean
support_matrix  <- OmegaCor[[1]]$support

name_map <- species_map %>%
  dplyr::filter(original_name_in_data %in% colnames(cor_matrix_mean)) %>%
  dplyr::select(original_name_in_data, pretty_name) %>%
  deframe()
new_labels <- name_map[colnames(cor_matrix_mean)]
colnames(cor_matrix_mean) <- rownames(cor_matrix_mean) <- new_labels
colnames(support_matrix)  <- rownames(support_matrix)  <- new_labels

supportLevel <- 0.95
p_matrix <- ifelse(support_matrix > supportLevel | support_matrix < (1 - supportLevel), 0.01, 1)

ggcorrplot::ggcorrplot(
  cor_matrix_mean, hc.order = TRUE, type = "lower",
  p.mat = p_matrix, insig = "blank", lab = FALSE,
  colors = c("#5AB4AC", "#F5F5F5", "#D8B365"),
  ggtheme = theme_minimal()
) +
  theme(axis.text.x = element_text(angle = 60, hjust = 1), axis.text.y = element_text(size = 8)) +
  labs(fill = "Correlation")
```

# 10. Future Climate Scenarios and Predictions (Sections 3.1.3, 3.2.3)

Species occurrence probabilities are predicted under four Shared
Socioeconomic Pathways (SSP1-2.6, SSP2-4.5, SSP3-7.0, SSP5-8.5) and three
30-year horizons (2011-2040, 2041-2070, 2071-2100), using downscaled
ClimateNA projections for MAT and MAP at each plot. Predictions use
`predictEtaMean = TRUE` (the posterior mean of the random-effect
realisations rather than a per-draw resample), a computationally efficient
simplification that slightly narrows -- but does not qualitatively change --
the uncertainty propagated into the Community Suitability Index below.

```{r future-climate-setup, eval = RUN_HEAVY}
fut_clim_raw <- read_csv(file.path(BASE_DIR, "climate", "plots_coord_12_GCMsY.csv"),
                          show_col_types = FALSE)
gcm_scenarios <- unique(fut_clim_raw$GCM)

XData_future_list <- list()
for (scenario in gcm_scenarios) {
  fut_clim_scenario <- fut_clim_raw %>%
    dplyr::filter(GCM == scenario) %>%
    dplyr::select(id1, MAT, MAP) %>%
    rename(plot_id = id1) %>%
    mutate(plot_id = as.character(plot_id))

  XData_future_list[[scenario]] <- XData %>%
    as_tibble(rownames = "plot_id") %>%
    left_join(fut_clim_scenario, by = "plot_id") %>%
    mutate(MAT = ifelse(is.na(MAT.y), MAT.x, MAT.y),
           MAP = ifelse(is.na(MAP.y), MAP.x, MAP.y)) %>%
    dplyr::select(-ends_with(".x"), -ends_with(".y")) %>%
    column_to_rownames("plot_id")
}
XData_future_list[["Current"]] <- XData
```

```{r future-climate-predict, eval = RUN_HEAVY}
# One posterior-prediction call per scenario; each returns a list of
# [sites x species] matrices, one per retained posterior draw.
for (scenario_name in names(XData_future_list)) {
  message("Predicting scenario: ", scenario_name)

  X_new <- XData_future_list[[scenario_name]]
  fac_cols <- names(Filter(is.factor, m1$XData))
  for (fc in fac_cols) {
    if (fc %in% colnames(X_new)) X_new[[fc]] <- factor(X_new[[fc]], levels = levels(m1$XData[[fc]]))
  }
  X_new <- X_new[, colnames(m1$XData), drop = FALSE]

  preds <- predict(m1, XData = X_new, expected = TRUE, predictEtaMean = TRUE)
  save(preds, file = file.path(DRAWS_DIR, paste0(scenario_name, "_preds.Rda")))
  rm(preds, X_new); gc()
}
```

```{r future-climate-load, eval = !RUN_HEAVY}
# --- Pre-computed posterior predictions per scenario -------------------------
# 13 files: "Current" (baseline) + 4 SSPs x 3 horizons, each a list of
# [sites x species] posterior-draw matrices, saved by the heavy chunk above.
scenario_files <- list.files(DRAWS_DIR, pattern = "_preds\\.Rda$", full.names = TRUE)
cat("Pre-computed prediction files found:", length(scenario_files), "\n")
print(basename(scenario_files))
```

# 11. Community Suitability Index (Section 3.3)

The Community Suitability Index (CSI) aggregates species-level occurrence
probabilities into a single measure of habitat suitability for a
pre-defined community. We compute the **geometric** CSI (gCSI), which
penalises communities where any single constituent species has low
suitability (a "weakest-link" property, consistent with Liebig's Law of the
Minimum), together with the basic (bCSI) and weighted (wCSI) arithmetic-mean
variants for completeness.

$$
gCSI_i = \left(\prod_{j=1}^{n} P_{ij}\right)^{1/n} = \exp\left(\frac{1}{n}\sum_{j=1}^{n}\ln(P_{ij})\right)
$$

Because HMSC yields a full posterior predictive distribution, we compute the
CSI **per posterior draw** and only then summarise across draws -- this
propagates parameter and prediction uncertainty consistently into the
reported credible intervals, rather than applying the (non-linear) geometric
mean to already-averaged summary statistics.

```{r csi-community-definitions}
# Ten representative forest community types (Table 1 of the manuscript),
# each defined by a fixed assemblage of 4-5 dominant tree species.
community_prefixes <- c("FE3", "FE6", "ME1", "MS6", "MS1", "MS2",
                         "RE2", "RC3", "RP1", "RS2")

type_eco_descriptions <- tibble::tribble(
  ~type_eco_prefix, ~Description,
  "FE3", "Sugar maple - Yellow birch Forest",
  "FE6", "Sugar maple - Red oak Forest",
  "ME1", "Black spruce - Trembling aspen Forest",
  "MS6", "Balsam fir - Red maple Forest",
  "MS1", "Balsam fir - Yellow birch Forest",
  "MS2", "Balsam fir - Paper birch Forest",
  "RE2", "Black spruce - Moss/heath Forest",
  "RC3", "Peaty eastern white cedar - Balsam fir Forest",
  "RP1", "White or red pine Forest",
  "RS2", "Balsam fir - Black spruce Forest"
)

type_eco_species_codes <- tibble::tribble(
  ~type_eco_prefix, ~dominant_codes, ~secondary_codes,
  "FE3", c("ERS", "BOJ"), c("HEG", "PRU", "SAB", "BOP", "FRA"),
  "FE6", c("ERS", "CHR"), c("BOJ", "BOP", "PIB", "ERR"),
  "ME1", c("EPN", "PET"), c("BOP", "SAB"),
  "MS6", c("SAB", "ERR"), c("BOP", "EPB"),
  "MS1", c("SAB", "BOJ"), c("ERS", "EPB"),
  "MS2", c("SAB", "BOP"), c("EPB", "EPN"),
  "RE2", c("EPN"),        c("SAB", "PIG", "BOP"),
  "RC3", c("THO", "SAB"), c("EPN", "MEL"),
  "RP1", c("PIB", "PIR"), c("ERR", "CHR", "BOP"),
  "RS2", c("SAB", "EPN"), c("BOP", "EPB")
)

community_defs <- purrr::pmap(
  list(type_eco_species_codes$type_eco_prefix,
       type_eco_species_codes$dominant_codes,
       type_eco_species_codes$secondary_codes),
  function(prefix, dom, sec) paste0(unique(c(dom, sec)), "_n")
)
names(community_defs) <- type_eco_species_codes$type_eco_prefix

knitr::kable(type_eco_descriptions, caption = "Table 1. The ten representative forest communities used for CSI computation.")
```

```{r csi-functions}
# --- Core CSI computation from posterior draws -------------------------------
# All three indices (gCSI, bCSI, wCSI) are computed per posterior draw and
# per site, then summarised (median, mean, SD, 95% credible interval) across
# draws -- see Section 3.3.4 of the manuscript for the full rationale.
epsilon <- 1e-4  # probability floor for numerical stability of log()

summarise_posterior <- function(x, central = "median") {
  x <- x[is.finite(x)]
  if (!length(x)) return(tibble(center = NA, sd = NA, q025 = NA, q975 = NA))
  tibble(
    center = if (central == "median") median(x) else mean(x),
    sd     = sd(x),
    q025   = quantile(x, 0.025, na.rm = TRUE),
    q975   = quantile(x, 0.975, na.rm = TRUE)
  )
}

coerce_pred_obj_to_draws <- function(pred_obj) {
  # 'preds' as returned by Hmsc::predict() is a list of [sites x species]
  # matrices, one per posterior draw; reshape to a [draw x site x species] array.
  n_draws <- length(pred_obj)
  site_names    <- rownames(pred_obj[[1]])
  species_names <- colnames(pred_obj[[1]])
  draws <- array(NA_real_, dim = c(n_draws, nrow(pred_obj[[1]]), ncol(pred_obj[[1]])),
                  dimnames = list(draw = paste0("draw_", seq_len(n_draws)),
                                  site = site_names, species = species_names))
  for (i in seq_len(n_draws)) draws[i, , ] <- pred_obj[[i]]
  draws
}

compute_gCSI_matrix <- function(draws_sub, eps = epsilon) {
  # Geometric mean in log-space, computed independently for every
  # (posterior draw x site) cell, across the species that define the community.
  logP <- log(pmax(draws_sub, eps))
  log_g_matrix <- apply(logP, c(1, 2), mean)
  list(g_matrix = exp(log_g_matrix), log_g_matrix = log_g_matrix)
}

summarise_community_gCSI <- function(log_g_matrix, central = "median") {
  # Community-level index: geometric mean of the site-level gCSI, i.e. the
  # mean of log(gCSI) across sites, taken independently for every draw, then
  # summarised across draws.
  comm_log <- apply(log_g_matrix, 1, function(row) mean(row[is.finite(row)]))
  g_stats  <- summarise_posterior(exp(comm_log), central = central)
  g_stats
}
```

```{r csi-run, eval = RUN_HEAVY}
scenario_map <- tibble::tribble(
  ~scenario_label,       ~projection,   ~file_name,
  "Current",             "Baseline",    "Current_preds.Rda",
  "SSP1-2.6 2011-2040",  "2011-2040",   "8GCMs_ensemble_ssp126_2011-2040.gcm_preds.Rda",
  "SSP1-2.6 2041-2070",  "2041-2070",   "8GCMs_ensemble_ssp126_2041-2070.gcm_preds.Rda",
  "SSP1-2.6 2071-2100",  "2071-2100",   "8GCMs_ensemble_ssp126_2071-2100.gcm_preds.Rda",
  "SSP2-4.5 2011-2040",  "2011-2040",   "8GCMs_ensemble_ssp245_2011-2040.gcm_preds.Rda",
  "SSP2-4.5 2041-2070",  "2041-2070",   "8GCMs_ensemble_ssp245_2041-2070.gcm_preds.Rda",
  "SSP2-4.5 2071-2100",  "2071-2100",   "8GCMs_ensemble_ssp245_2071-2100.gcm_preds.Rda",
  "SSP3-7.0 2011-2040",  "2011-2040",   "8GCMs_ensemble_ssp370_2011-2040.gcm_preds.Rda",
  "SSP3-7.0 2041-2070",  "2041-2070",   "8GCMs_ensemble_ssp370_2041-2070.gcm_preds.Rda",
  "SSP3-7.0 2071-2100",  "2071-2100",   "8GCMs_ensemble_ssp370_2071-2100.gcm_preds.Rda",
  "SSP5-8.5 2011-2040",  "2011-2040",   "8GCMs_ensemble_ssp585_2011-2040.gcm_preds.Rda",
  "SSP5-8.5 2041-2070",  "2041-2070",   "8GCMs_ensemble_ssp585_2041-2070.gcm_preds.Rda",
  "SSP5-8.5 2071-2100",  "2071-2100",   "8GCMs_ensemble_ssp585_2071-2100.gcm_preds.Rda"
) %>% mutate(file_path = file.path(DRAWS_DIR, file_name))

community_results <- list()
for (i in seq_len(nrow(scenario_map))) {
  scn <- scenario_map[i, ]
  env <- new.env(parent = emptyenv())
  load(scn$file_path, envir = env)
  draws_array <- coerce_pred_obj_to_draws(env$preds)
  available_species <- dimnames(draws_array)[[3]]

  for (prefix in names(community_defs)) {
    species_vec <- intersect(community_defs[[prefix]], available_species)
    if (!length(species_vec)) next
    species_idx <- match(species_vec, available_species)
    g_list  <- compute_gCSI_matrix(draws_array[, , species_idx, drop = FALSE])
    comm_g  <- summarise_community_gCSI(g_list$log_g_matrix)

    community_results[[length(community_results) + 1]] <- comm_g %>%
      mutate(scenario_label = scn$scenario_label,
             projection_horizon = scn$projection,
             community_prefix = prefix)
  }
}

trajectory_results <- bind_rows(community_results) %>%
  rename(gCenter = center, gSD = sd, gQ025 = q025, gQ975 = q975) %>%
  left_join(type_eco_descriptions, by = "type_eco_prefix" = "community_prefix")

write_csv(trajectory_results, file.path(OUTPUTS_DIR, "gCSI_trajectory_summary_draws.csv"))
```

```{r csi-load, eval = !RUN_HEAVY}
# --- Load the pre-computed, per-draw CSI summary -----------------------------
trajectory_results <- read_csv(file.path(OUTPUTS_DIR, "gCSI_trajectory_summary_draws.csv"),
                                show_col_types = FALSE)
```

# 12. Community Suitability Trajectories Under Climate Change (Figure 3)

```{r csi-trajectories-prep}
trajectory_results <- trajectory_results %>%
  mutate(
    scenario_label = factor(scenario_label, levels = c(
      "Current",
      "SSP1-2.6 2011-2040", "SSP1-2.6 2041-2070", "SSP1-2.6 2071-2100",
      "SSP2-4.5 2011-2040", "SSP2-4.5 2041-2070", "SSP2-4.5 2071-2100",
      "SSP3-7.0 2011-2040", "SSP3-7.0 2041-2070", "SSP3-7.0 2071-2100",
      "SSP5-8.5 2011-2040", "SSP5-8.5 2041-2070", "SSP5-8.5 2071-2100"
    )),
    community_label = factor(community_prefix, levels = community_prefixes)
  ) %>%
  left_join(type_eco_descriptions, by = c("community_prefix" = "type_eco_prefix"))
```

```{r csi-trajectories-table}
trajectory_results %>%
  dplyr::filter(scenario_label %in% c("Current", "SSP3-7.0 2071-2100")) %>%
  dplyr::select(Description, scenario_label, gCenter, gQ025, gQ975) %>%
  arrange(Description, scenario_label) %>%
  knitr::kable(digits = 3,
               col.names = c("Community", "Scenario", "Median gCSI", "2.5%", "97.5%"),
               caption = "Geometric Community Suitability Index (gCSI): current baseline vs. SSP3-7.0, 2071-2100.")
```

```{r csi-trajectories-plot, fig.width = 9, fig.height = 8, fig.cap = "Projected community-level suitability (gCSI) under four SSP scenarios and three future horizons, relative to the current baseline (dashed line)."}
ggplot(trajectory_results,
       aes(x = scenario_label, y = gCenter,
           colour = scenario_label, group = Description)) +
  geom_hline(data = trajectory_results %>% dplyr::filter(scenario_label == "Current"),
             aes(yintercept = gCenter), linetype = "dashed", colour = "grey50") +
  geom_errorbar(aes(ymin = gQ025, ymax = gQ975), width = 0, linewidth = 0.6) +
  geom_point(size = 2) +
  facet_wrap(~ Description, scales = "free_y", ncol = 2) +
  labs(x = NULL, y = "Geometric CSI (gCSI)", colour = "Scenario") +
  theme_minimal(base_size = 10) +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        legend.position = "bottom")
```

# 13. Appendix: Preparing Posterior-Draw Subsamples for the TCAMS App

To keep the Tree Community-Assisted Migration Simulator (TCAMS, Section 3.5)
responsive, the full posterior-draw prediction files (thousands of draws per
scenario) are subsampled down to a manageable number of draws (`K`) before
being bundled with the Shiny application. This is a one-off data-preparation
utility, not part of the statistical analysis itself, and is therefore shown
here for completeness but not executed.

```{r tcams-subsample, eval = FALSE}
original_draws_dir   <- file.path(RESULTS_DIR, "preds")
subsampled_draws_dir <- file.path(BASE_DIR, "posterior_draws_200")
K <- 400L

load_preds <- function(path) {
  env <- new.env(parent = emptyenv())
  load(path, envir = env)
  env$preds
}

subsample_preds <- function(preds, K) {
  n_draws <- length(preds)
  keep_n  <- min(K, n_draws)
  idx <- if (keep_n < n_draws) sort(sample.int(n_draws, keep_n)) else seq_len(n_draws)
  preds[idx]
}

dir.create(subsampled_draws_dir, recursive = TRUE, showWarnings = FALSE)
rda_files <- list.files(original_draws_dir, pattern = "_preds\\.Rda$", full.names = TRUE)

for (rda_path in rda_files) {
  scenario_label <- sub("_preds\\.Rda$", "", basename(rda_path))
  preds_full <- load_preds(rda_path)
  preds <- subsample_preds(preds_full, K)
  save(preds, file = file.path(subsampled_draws_dir,
                                sprintf("%s_preds_subsampled.Rda", scenario_label)))
  message(sprintf("%s: kept %d of %d draws", scenario_label, length(preds), length(preds_full)))
}
```

# Session Information

```{r session-info}
sessionInfo()
```
